Followers Are Not Enough: A Question-Oriented Approach to Community Detection in Online Social Networks

### 2014-08-18 paper-for_arxiv/methods.tex
minor edits

This last criterion in particular poses a problem because the majority of community detection algorithms developed so far are built to find partitions of a network and few are aimed at finding overlapping communities \cite{Aicher26062014,BaumesGKMP05,PalEtAl05,ZhaWanZha07,Gre07,PhysRevE.77.016107,Lancichinetti2009,PhysRevE.80.016105,Kovacs2010}. Among these methods, even fewer deal with directed or weighted networks. For example, the work of~\cite{PalEtAl05} on clique percolation can account for both features, but not at the same time. A recent method proposed by~\cite{LancichinettiPlos}, OSLOM (Order Statistics Local Optimization Method), is one of the first methods  able to deal with all of these features simultaneously. Their method relies on a fitness function that measures the statistical significance of clusters with respect to random fluctuations, and attempts to optimize this fitness function across all clusters. 
Following~\cite{LancichinettiPlos}, the significance measure is defined as the probability of finding the cluster in a network without community structure. The null model used is basically the same as the one adopted by Newman and Girvan to define modularity, i.e. it is a model that generates random graphs with a given degree distribution. The authors tested their algorithm on different benchmarks (LFR and Girvan-Newman) and real networks (such as the US air transportation and the word association network), and compared its performance against the best algorithms
currently available (i.e the ones mentioned in the introduction), and found excellent results. Moreover, they showed that OSLOM is also able to recognize the absence, and not simply the presence, of community structure, by testing it on random graphs. This feature of the algorithm plays an important role in the analysis of real data networks, since it is not always the case that a community structure is indeed present and it is therefore useful to be able to detect its absence too. 
Therefore, given its versatility and performance with benchmark network, in this paper we used OSLOM to detect \emph{overlapping} communities present in our \emph{weighted} and \emph{directed} networks.
%For this paper we use OSLOM since it allowed us to detect \emph{overlapping} communities present in our \emph{weighted} and \emph{directed} network, and its authors showed that it performs very well on various types of artificial benchmark graphs and also on several real networks.

@@

Since our goal is to explore the functional communities of this network, we filter the network down to the subset of users which actively interact with each other (e.g., via retweets and mentions). We do this by measuring what we call (incoming/outgoing) information events. We define an outgoing information event for a given user $u$ as either a mention made by $u$ of another user in the network, or a retweet of one of $u$'s tweets by another user in the network. The logic for this definition is as follows: if $u$ mentions a user $v$ this can be thought of as $u$ directly sending information to $v$, and if $u$ is retweeted by $v$ then $v$ received information from $u$ and rebroadcast it to their followers. In either case there was information outgoing from $u$ which affected the network in the some way. Analogously, we define the incoming information event for $u$ as either being mentioned by a different user in the network, or as retweeting another user in the network.
%function and information share is important We then filter down this starting %network to include only users that [[what??]]. To do this 
%Since one of the community types we want to explore is based on the explicit interactions ( between users, we filter the dataset in order to take into account only users who make use of these kinds of features in their tweets. [[Seeems out of place]]
%We define an event of outgoing information for a given user $u$ as either a mention made by $u$ of another user in the network, or a retweet by another user of one of $u$'s tweets. 
%[[Switched the readers perspective]]When we mention someone we are in fact sending him some information, and when we are retweeted it means that the person that retweeted us has received some information from us and is sharing it. 
With (incoming/outgoing) information events defined we filtered the network by eliminating all users with less than a total of 9 outgoing \emph{and} incoming information events, i.e., less than one information event per type per week on average. 
We then further restricted our analysis to the strong giant connected component of the network built from the (incoming/outgoing) information filtered set of users. %and whose link represent a user-follower relationship.
In this study the link is directed from the user to the follower because this is the direction in which the information (in the case of transfer entropy) or influence (in the case of mention-retweets) flows. Thus, for a pair of users $u$ and $v$, an edge $a_{v \to u}$ in the structural network has weight 1 if user $u$ follows $v$, and 0 otherwise. The final network consists of 6,917 nodes and 1,481,131 edges.

@@

Operationally, we expect users to interact with Twitter on a human time scale, and thus the natural one-second time resolution is too fine since most humans do not write tweets on the time scale of seconds. We coarsen each time series by considering non-overlapping time intervals ten minutes in length. For each time interval, we record a 1 if the user has tweeted during that time interval, and a 0 if they have not. Thus, the new coarsened time series now captures whether or not the user has been active on Twitter over any given ten minute time interval in our data set. We can then compute the flow of information between two users $u$ and $v$ by computing the transfer entropy between their time series $X_{t}(u)$ and $X_{t}(v).$ See Appendix A for a detailed introduction to transfer entropy and its estimation from data.

@@

 We computed the transfer entropy on each coarsened time series with lag $k$ ranging from 1 to 6, this corresponds to a lag of ten minutes to an hour. The choice of lag must balance a trade-off between additional information and sparsity of samples: as we increase the lag $k$, we account for longer range dependencies, but we also decrease the number of samples available to infer a higher dimensional predictive distribution. See Appendix A for a discussion of the estimation of transfer entropy from data. As we will see, the underlying communities resulting from the different lags have similar structure.

@@

The final community we consider is a topic-based or topical community, i.e., a community defined by the content (topics) users discuss. So in a topical community, users are defined to be a member of a community if they tweet \emph{about} similar topics as other members of the community. 
%Another kind of community is the one based on the content of the tweets, and relies on the idea of finding people that talk about the same things. 
In order to detect the topical communities, we weight the edges of the user-follower network through a measure based on the number of common hashtags between pairs of users. Hashtags are a good proxy for a tweet's content as hashtags are explicitly meant to be keywords indicating the topic of the tweet. Moreover they are widely used and straightforward to detect.

To this end, we characterize each user $u$ by a vector $\vec{h}(u)$ of length equal to the number of unique hashtags in the dataset, and whose elements are defined as
\begin{equation}
h_i(u) = n_i(u)\log{ \frac{N}{n_i} }
\end{equation}
where $n_i(u)$ is the frequency of hashtag $i$ occuring in user $u$'s tweets, $N$ is the total number of users, and $n_i$ is the number of users that have used the hashtag $i$ in their tweets. This adapted term frequency--inverse document frequency (tf-idf) measure \cite{salton_introduction_1983} captures the importance of a hashtag in the users's tweets through the first factor, but at the same time smooths it through the second factor by giving less importance to hashtags that are too widely used (as $\frac{N}{n_i}$ approaches one, its logarithm approaches zero).

>>>>

This last criterion in particular poses a problem because the majority of community detection algorithms developed so far are built to find partitions of a network and few are aimed at finding overlapping communities \cite{Aicher26062014,BaumesGKMP05,PalEtAl05,ZhaWanZha07,Gre07,PhysRevE.77.016107,Lancichinetti2009,PhysRevE.80.016105,Kovacs2010}. Among these methods, even fewer deal with directed or weighted networks. For example, the work of~\cite{PalEtAl05} on clique percolation can account for both features, but not at the same time. A recent method proposed in~\cite{LancichinettiPlos}, OSLOM (Order Statistics Local Optimization Method), is one of the first methods  able to deal with all of these features simultaneously. This method of Lancichinetti et al. relies on a fitness function that measures the statistical significance of clusters with respect to random fluctuations, and attempts to optimize this fitness function across all clusters. 
Following~\cite{LancichinettiPlos}, the significance measure is defined as the probability of finding the cluster in a network without community structure. The null model used is basically the same as the one adopted by Newman and Girvan in \cite{newman2004finding} to define modularity, {\it i.e.} it is a model that generates random graphs with a given degree distribution. The authors tested their algorithm on different benchmarks (LFR and Girvan-Newman) and real networks (such as the US air transportation and the word association network), and compared its performance against the best algorithms
currently available (i.e the ones mentioned in the introduction), and found excellent results. Moreover, they showed that OSLOM is also able to recognize the absence, and not simply the presence, of community structure, by testing it on random graphs. This feature of the algorithm plays an important role in the analysis of real data networks, since it is not always the case that a community structure is indeed present and it is therefore useful to be able to detect its absence too. 
Therefore, given its versatility and performance with benchmark networks, in this paper we used OSLOM to detect \emph{overlapping} communities present in our \emph{weighted} and \emph{directed} networks.
%For this paper we use OSLOM since it allowed us to detect \emph{overlapping} communities present in our \emph{weighted} and \emph{directed} network, and its authors showed that it performs very well on various types of artificial benchmark graphs and also on several real networks.

@@

Since our goal is to explore the functional communities of this network, we filter the network down to the subset of users which actively interact with each other (e.g., via retweets and mentions). We do this by measuring what we call (incoming/outgoing) information events. We define an outgoing information event for a given user $u$ as either a mention made by $u$ of another user in the network, or a retweet of one of $u$'s tweets by another user in the network. The logic for this definition is as follows: if $u$ mentions a user $v$ this can be thought of as $u$ directly sending information to $v$, and if $u$ is retweeted by $v$ then $v$ received information from $u$ and rebroadcast it to their followers. In either case there was information outgoing from $u$ which affected the network in some way. Analogously, we define the incoming information event for $u$ as either being mentioned by a different user in the network, or as retweeting another user in the network.
%function and information share is important We then filter down this starting %network to include only users that [[what??]]. To do this 
%Since one of the community types we want to explore is based on the explicit interactions ( between users, we filter the dataset in order to take into account only users who make use of these kinds of features in their tweets. [[Seeems out of place]]
%We define an event of outgoing information for a given user $u$ as either a mention made by $u$ of another user in the network, or a retweet by another user of one of $u$'s tweets. 
%[[Switched the readers perspective]]When we mention someone we are in fact sending him some information, and when we are retweeted it means that the person that retweeted us has received some information from us and is sharing it. 
With (incoming/outgoing) information events defined we filtered the network by eliminating all users with less than a total of 9 outgoing \emph{and} incoming information events, {\it i.e.,} less than one information event per type per week on average. 
We then further restricted our analysis to the strong giant connected component of the network built from the (incoming/outgoing) information filtered set of users. %and whose link represent a user-follower relationship.
In this study the link is directed from the user to the follower because this is the direction in which information flows. Thus, for a pair of users $u$ and $v$, an edge $a_{v \to u}$ in the structural network has weight 1 if user $u$ follows $v$, and 0 otherwise. The final network consists of 6,917 nodes and 1,481,131 edges.

@@

Operationally, we expect users to interact with Twitter on a human time scale, and thus the natural one-second time resolution is too fine; since most humans do not write tweets on the time scale of seconds. For this reason, we coarsen each time series by considering non-overlapping time intervals ten minutes in length. For each time interval, we record a 1 if the user has tweeted during that time interval, and a 0 if they have not. Thus, the new coarsened time series now captures whether or not the user has been active on Twitter over any given ten minute time interval in our data set. We can then compute the flow of information between two users $u$ and $v$ by computing the transfer entropy between their time series $X_{t}(u)$ and $X_{t}(v).$ See Appendix A for a detailed introduction to transfer entropy and its estimation from data.

@@

 We computed the transfer entropy on each coarsened time series with lag $k$ ranging from 1 to 6, this corresponds to a lag of ten minutes to an hour. The choice of lag must balance a trade-off between additional information and sparsity of samples: as we increase the lag $k$, we account for longer range dependencies, but we also decrease the number of samples available to infer a higher dimensional predictive distribution. See Appendix A for a discussion of these issues. As we will show, the underlying communities resulting from the different lags have similar structure.

@@

The final community we consider is a topic-based or topical community, i.e., a community defined by the content (topics) users discuss. In a topical community, users are defined to be a member of a community if they tweet \emph{about} similar topics as other members of the community. 
%Another kind of community is the one based on the content of the tweets, and relies on the idea of finding people that talk about the same things. 
In order to detect the topical communities, we weight the edges of the user-follower network through a measure based on the number of common hashtags between pairs of users. Hashtags are a good proxy for a tweet's content as hashtags are explicitly meant to be keywords indicating the topic of the tweet. Moreover they are widely used and straightforward to detect.

To this end, we characterize each user $u$ by a vector $\vec{h}(u)$ of length equal to the number of unique hashtags in the dataset, and whose elements are defined as
\begin{equation}
h_i(u) = \phi_i(u)\log{ \frac{N}{n_i} }
\end{equation}
where $\phi_i(u)$ is the frequency of hashtag $i$ occuring in user $u$'s tweets, $N$ is the total number of users, and $n_i$ is the number of users that have used the hashtag $i$ in their tweets. This adapted term frequency--inverse document frequency (tf-idf) measure \cite{salton_introduction_1983} captures the importance of a hashtag in the users's tweets through the first factor, but at the same time smooths it through the second factor by giving less importance to hashtags that are too widely used (as $\frac{N}{n_i}$ approaches one, its logarithm approaches zero).

### 2014-08-18 paper-for_arxiv/paper.tex
Removed DARPA acknowledgement.
formatting
minor edits

\section{Funding}
%Elisa Omodei is partially supported by a PhD grant from the R\'{e}gion \^{I}le-de-France (DIM 2011) and by the French National Agency of Research (ANR) through the grant Algopol (ANR-12-CORD-0018).
This work was supported by the R\'{e}gion \^{I}le-de-France [DIM 2011 to E.O.]; and by the French National Agency of Research [ANR-12-CORD-0018 to E.O.].
We are grateful to DARPA for the support of David Darmon.

>>>>

\section{Funding}
%Elisa Omodei is partially supported by a PhD grant from the R\'{e}gion \^{I}le-de-France (DIM 2011) and by the French National Agency of Research (ANR) through the grant Algopol (ANR-12-CORD-0018).
This work was partially supported by the R\'{e}gion \^{I}le-de-France [DIM 2011 to E.O.]; and by the French National Agency of Research [ANR-12-CORD-0018 to E.O.].
%We are grateful to DARPA for the support of David Darmon.

### 2014-08-18 paper-for_arxiv/results.tex
Modified ecdf figure to match LaTeX font.
minor edits

\begin{figure}[ht]
  \centering
%\includegraphics[width=0.50\textwidth]{comm_sizes_ecdf_loglog.pdf}
\includegraphics[width=0.6\textwidth,angle=270]{cdf_comm_sizes.eps}
\caption{The proportion of communities greater than $s$ in size, across the different community types. Note the logarithmic scale on the horizontal and vertical axes.}
\label{Fig-community_size_distribution}
\end{figure}

@@

Any covering determined by OSLOM induces a natural partition of the edges in a directed network. In particular, Let $u$ and $f$ be two users in the network, and let $M_{u}$ and $M_{f}$ be their community memberships. Then any edge $e_{u \to f}$ can be partitioned into one of three classes by
\begin{align}
	T(e_{u \to f}) = \left\{ \begin{array}{ll}
		\text{Inter-edge} &: \ \ \ M_{u} \cap M_{f} = \emptyset \\
		\text{Intra-edge} &: \ \ \ M_{u} = M_{f}\\
		\text{Mixed-edge} &: \ \ \ \text{ otherwise} \\
	\end{array}\right. \label{Eqn-edge_types}.
\end{align}
In words, an inter-edge connects two users who share no community memberships, an intra-edge connects two users who each belong to the same communities, and a mixed-edge connects two users who share some, but not all, community memberships. Thus, inter-edges cross community boundaries, intra-edges lie within community boundaries, and mixed-edges lie at the borders of community boundaries. See Figure~\ref{Fig-edge_types} for a schematic of this edge partitioning.

>>>>

\begin{figure}[ht]
  \centering
%\includegraphics[width=0.50\textwidth]{comm_sizes_ecdf_loglog.pdf}
\includegraphics[width=1\textwidth]{cdf_comm_sizes.pdf}
\caption{The proportion of communities greater than $s$ in size, across the different community types. Note the logarithmic scale on the horizontal and vertical axes.}
\label{Fig-community_size_distribution}
\end{figure}

@@

Any covering determined by OSLOM induces a natural partition of the edges in a directed network. In particular, let $u$ and $f$ be two users in the network, and let $M_{u}$ and $M_{f}$ be their community memberships. Then any edge $e_{u \to f}$ can be partitioned into one of three classes by
\begin{align}
	T(e_{u \to f}) = \left\{ \begin{array}{ll}
		\text{Inter-edge} &: \ \ \ M_{u} \cap M_{f} = \emptyset \\
		\text{Intra-edge} &: \ \ \ M_{u} = M_{f}\\
		\text{Mixed-edge} &: \ \ \ \text{ otherwise} \\
	\end{array}\right. \label{Eqn-edge_types}.
\end{align}
In words, an inter-edge connects two users who share no community memberships, an intra-edge connects two users who each belong to the same communities, and a mixed-edge connects two users who share some, but not all, community memberships. Thus, inter-edges cross community boundaries, intra-edges lie within community boundaries, and mixed-edges lie at the borders of community boundaries. See Figure~\ref{Fig-edge_types} for a schematic of this edge partitioning.
